---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  fig.path = "tools/readme/"
)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
devtools::load_all()
set.seed(0)
```

# pheatbuilder

This package provides a stepwise, pipe-friendly interface to create heatmaps in
R.

## Installation

You can install pheatbuilder from github with:

```{r eval=FALSE}
# install.packages("devtools")
devtools::install_github("kylebittinger/pheatbuilder")
```

## Making heatmaps with pheatbuilder

The pheatbuilder package comes with a built-in dataset, `vendor_props`. This
data matrix contains bacterial taxon proportions in cecal and fecal samples
in mice from four different vendors.

Heatmaps start by passing a matrix or data frame to the `pheat` function. By
default, the cells of the heatmap are 10pt square (matching the default font
size), and the rows/columns are not clustered.

```{r}
pheat(vendor_props)
```

To adjust the heatmap, additional functions can be chained. Let's cluster
the rows and remove the cell borders.

```{r warning=FALSE, message=FALSE}
library(magrittr)
vendor_props %>%
  pheat() %>%
  pheat_cluster_rows() %>%
  pheat_display_main(border_color = NA)
```

If you have data frames with additional info for the rows and columns, thse can
be added as annotations to the heatmap. The data frame `vendor_samples` has
info about the sample types and vendors.

```{r}
head(vendor_samples)
```

By default, values in the first column are used to match to the heatmap grid.

```{r}
vendor_props %>%
  pheat() %>%
  pheat_annotate_cols(vendor_samples)
```

Heatmaps can be saved with a call to `pheat_save`. If the height and width
are not set, the document is automatically sized to fit the heatmap.

```{r eval=FALSE}
vendor_props %>%
  pheat() %>%
  pheat_save("vendor_heatmap.pdf")
```


The function `pheat_save` returns the heatmap invisibly. If you still want to
see the heatmap on screen after it is saved to a file, you can add `print` to
the chain of functions at the end.

## Convenience functions for gaps and color palettes

Sometimes it's nice to have gaps between rows or columns of the heatmap. The
gap locations are typically specified with row or column numbers, after which
the gaps are to appear. As a convenience, we provide the `factor_gaps` function
to generate gap locations automatically based on a factor or character vector.

For example, we can create a vector of gap locations for the mouse vendor.

```{r}
factor_gaps(vendor_samples$vendor)
```

And here is a vector of gaps between bacterial phyla.

```{r}
factor_gaps(vendor_taxa$phylum)
```

Putting these both in the plot, we can add gaps in the rows and columns.

```{r}
vendor_props %>%
  pheat() %>%
  pheat_display_cols(gaps = factor_gaps(vendor_samples$vendor)) %>%
  pheat_display_rows(gaps = factor_gaps(vendor_taxa$phylum))
```

One trouble spot with annotated heatmaps is setting the color palettes for
various annotations. The function `factor_palette` allows the user to create a
named vector of colors, and to pull specific colors to the front if needed.

Here, we take the third color from the "Set 2" palette and use it as the first
color in our palette for sample types. The remaining colors in the "Set 2"
palette are used as needed for the additional sample types.

For vendors, we use only the odd colors from the "Paired" palette, to keep the
color scheme light.

```{r}
sample_type_colors <- factor_palette(
  vendor_samples$sample_type, 
  palette.colors(palette = "Set 2"), 
  3)
vendor_colors <- factor_palette(
  vendor_samples$vendor,
  palette.colors(palette = "Paired"),
  1, 3, 5, 7)
vendor_props %>%
  pheat() %>%
  pheat_annotate_cols(vendor_samples) %>%
  pheat_annotation_color(
    sample_type = sample_type_colors,
    vendor = vendor_colors)
```

Another convenience function is `pheat_color_saturated`, which applies a
saturated-rainbow color palette to the heatmap. It's ugly, but it works to
highlight values that are zero (white) or above 0.4 (red).

```{r}
vendor_props %>%
  pheat() %>%
  pheat_color_saturated(saturation_limit = 0.4)
```

